A statistically consistent approach to chemical tagging

### 2018-12-16 article/ms.tex
moo

\begin{figure*}
	\includegraphics[width=1.0\textwidth]{experiments/toy-model-data.png}
    \caption{A corner plot showing all $\NumDimensions$ dimensions of the
    		 generated data for Experiment~1 (blue). Overplotted in red
		 	 we show samples generated given our estimate of the model 
			 parameters, without adding specific variance in each dimension.
			 This demonstrates that the model parameters estimated by
			 expectation-maximization can replicate the generated data.}
    \label{fig:toy-model-data}
\end{figure*}

\begin{figure*}
	\includegraphics[width=1.0\textwidth]{experiments/toy-model-latent-space.png}
    \caption{A corner plot showing all $\NumLatentFactors$ dimensions in
    		 latent space, where each data point is coloured by the component
		 	 that it is estimated to belong to.}
    \label{fig:toy-model-latent-space}
\end{figure*}

\subsubsection{Model fitting given the true number of latent factors and components}

@@

\begin{figure}
	%\includegraphics[width=0.5\textwidth]{experiments/toy-model-data.png}
    \caption{Log-likelihood of the model given the data $\mathcal{L}\left(\vec\data|\vec\Psi\right)$ for Experiment 1 as a function of the number of expectation-maximization iterations. After \todo{X} iterations the improvement in log-likelihood per
    expectation-maximization cycle was less than $10^{-5}$.}
    \label{fig:toy-log-likelihood-per-iteration}
\end{figure}

@@

While the estimated model parameters can generate the data and demonstrate
clustering in latent space, caution is needed when interpreting the 
model latent variables.
Latent factor models are by definition indeterminate in that there are
multiple solutions of the parameter vector $\vec\Psi$ which provide
an identical data-generating process. This can be trivially shown by
the fact that the factors loads and scores can be rotated by taking the
dot product
with some rotation
matrix $\textbf{R}$ and generate the exact same data. As a consequence,
we cannot strictly evaluate our model fitting procedure by comparing
how closely the inferred factor loads (or scores) match the true 
variables. This is further complicated by orientation and order
of the latent factors: even if the inferred factor loads were very close
to the true factor loads, they could be ordered differently, or have
their signs reversed. We discuss this further in later sections.

\subsubsection{A grid search for the number of latent factors and components}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/toy-ll-contours.png}
	\includegraphics[width=0.45\textwidth]{experiments/toy-bic-contours.png}
	\includegraphics[width=0.45\textwidth]{experiments/toy-pseudobic-contours.png}
    \caption{Model performance metrics resulting from a gridsearch using the
    		 data generated as part of our toy model. The top 
		 	 panel shows the negative log-likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors $J$ and number 
			 of clusters $K$, the middle panel shows the BIC (Equation \ref{eq:bic}), 
			 and the lower panel shows
			 the pseudo-BIC
			 (Equation~\ref{eq:pseudo-bic}). The marker indicates the 
			 lowest value in each panel.}
    \label{fig:experiment-1-gridsearch}
\end{figure}

@@

The estimated latent factor loads found during this grid search are shown in
Figure~\ref{fig:fig:experiment-1-gridsearch-factors-init-random-and-random},
where each panel shows the estimated latent factor loads found by
expectation-maximization for a given number of latent factors and components.
The latent factors $\factorloads$ in each panel are multiplied by a rotation
matrix $\mathbf{R}$ with entries $\{-1, 0, 1\}$ in order to re-orient and flip
the estimated latent factors to be as similar as possible to the latent factors
found from the model with the same number of latent factors and $\NumComponents  =  1$.
In doing so we find that the estimated latent factors vary slightly as
the number of components $\NumComponents$ increases, but the overall shape
of the latent factors remains the same.

\begin{figure*}
	\includegraphics[width=1.0\textwidth]{experiments/toy-factors-init-randomly.png}
	\caption{Latent factor loads inferred for models with increasing numbers
			 of latent factors and increasing numbers of cluster components
			 in Experiment~1, where the latent factors and component assignments
			 were initialised randomly.
				The top left panel shows the results of the model with
			 $\NumLatentFactors = 1$ and $\NumComponents = 1$. 
			 The latent factors in each panel have been rotated by a
			 rotation matrix $\textbf{R}$ with entries $\{-1, 0, +1\}$ such 
			 that the estimated latent factors in each model are ordered 
			 and oriented to be as similar as the latent factors estimated from
			 the model with only one component (i.e., the latent factors in the top row
			 of the same column). Note that the rotation
			 matrix contains only entries $\{-1, 0, +1\}$.}
	\label{fig:toy-factors-init-randomly}
\end{figure*}

\todo{This structure goes away and becomes more consistent if we change the initialisation
procedure}

\todo{figure of toy factors init svd kmeans++}

\todo{log-likelihood of different initialisation routines}

\begin{figure*}
	\includegraphics[width=1.0\textwidth]{experiments/toy-svd-kmeans-factors.png}
	\caption{Latent factor loads inferred for models with increasing numbers
			 of latent factors and increasing numbers of cluster components
			 in Experiment~1, where the latent factors were initialised using
			 \texttt{SVD} and the component assignments were initialised using \texttt{k-means++}.
				The top left panel shows the results of the model with
			 $\NumLatentFactors = 1$ and $\NumComponents = 1$. 
			 The latent factors in each panel have been rotated by a
			 rotation matrix $\textbf{R}$ with entries $\{-1, 0, +1\}$ such 
			 that the estimated latent factors in each model are ordered 
			 and oriented to be as similar as the latent factors estimated from
			 the model with only one component (i.e., the latent factors in the top row
			 of the same column). Compared to Figure~\ref{fig:toy-factors-init-randomly},
			 it is clear that initialising with \texttt{SVD}/\texttt{k-means++} results
			 in less scatter in the final estimates of the model parameters.}
	\label{fig:toy-factors-init-svd-and-kmeans++}
\end{figure*}

The results of this experiment confirm that the model is indeterminate, in 
that there are multiple equally valid solutions that can exactly reproduce
the data. It also highlights the importance of initialisation, demonstrating
how the initialisation of the common latent factors can impact the final
model parameters. These two points are related, as one could initialise the
latent factors randomly (or from noise) and multiply that model with a
rotation matrix $\mathbf{R}$ that rotates the latent space such that the
common factor loads are aligned closer to some prior expectation that we have
for the common factor loads (e.g., in that they should represent plausible
astrophysical yields). Motivated by these results and indeterminate nature
of the model, in Experiment~2 we initialise the common factor loads
to represent known astrophysical processes such that they can be interpreted
in context with the data.

@@

\todo{experiment-4-astrophysical-latent-factors}

\subsection{Experiment 3: A subset of gravitationally bound stellar clusters from \Galah\ and elsewhere}
\label{sec:experiment-galah-clusters}

@@

\begin{figure*}
	\includegraphics[width=1.0\textwidth]{experiments/galah-experiment-2-gridsearch-factors-wrt-J-and-K.png}
	\caption{Latent factor loads inferred for models with increasing numbers
			 of latent factors and increasing numbers of cluster components
			 in Experiment~2. The latent factors have each been rotated by a
			 rotation matrix $\textbf{R}$ with entries $\{-1, 0, +1\}$ such 
			 that the estimated latent factors in each model are ordered 
			 and oriented to be as similar as the latent factors estimated from
			 the model with only one latent factor. Note that the rotation
			 matrix contains only entries $\{-1, 0, +1\}$. The latent
			 factors for each model shown here were initialised randomly.
			 }
	\label{fig:experiment-2-gridsearch-factors-init-svd}
\end{figure*}

>>>>

\begin{figure*}
	\includegraphics[width=1.0\textwidth]{experiments/exp1-data.png}
    \caption{A corner plot showing all $\NumDimensions$ dimensions of the
    		 generated data for Experiment~1 (blue). Overplotted in red
		 	 we show samples generated given our estimate of the model 
			 parameters, without adding specific variance in each dimension.
			 This demonstrates that the model parameters estimated by
			 expectation-maximization can replicate the generated data.}
    \label{fig:exp1-data}
\end{figure*}

%\begin{figure*}
%	\includegraphics[width=1.0\textwidth]{experiments/toy-model-latent-space.png}
%    \caption{A corner plot showing all $\NumLatentFactors$ dimensions in
%    		 latent space, where each data point is coloured by the component
%		 	 that it is estimated to belong to.}
%    \label{fig:toy-model-latent-space}
%\end{figure*}

\subsubsection{Given the true number of latent factors and components}

@@

\begin{figure}
	\includegraphics[width=0.5\textwidth]{experiments/exp1-ll-iterations.png}
	\caption{Log-likelihood of the model given the data $\mathcal{L}\left(\vec\data|\vec\Psi\right)$ for Experiment 1 as a function of the number of expectation-maximization iterations. After \todo{X} iterations the improvement in log-likelihood per
    expectation-maximization cycle was less than $10^{-5}$.}
    \label{fig:exp1-ll-iterations.png}
\end{figure}

@@

While the estimated model parameters can generate the data and demonstrate
clustering in latent space, caution is needed when interpreting the 
model latent variables.
Latent factor models are by definition indeterminate in that there are
multiple solutions of the parameter vector $\vec\Psi$ which provide
an identical data-generating process. This can be trivially shown by
the fact that the factors loads and scores can be rotated by taking the
dot product
with some rotation
matrix $\textbf{R}$ and generate the exact same data. 
We discuss how this limitation applies to later experiments.

%This is further complicated by orientation and order
%of the latent factors: even if the inferred factor loads were very close
%to the true factor loads, they could be ordered differently, or have
%their signs reversed. We discuss this further in later sections.

\subsubsection{Grid search for the number of latent factors and components}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-ll-contours.png}
	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-bic-contours.png}
%	\includegraphics[width=0.45\textwidth]{experiments/toy-pseudobic-contours.png}
    \caption{Model performance metrics resulting from a gridsearch using the
    		 data generated as part of our toy model. The top 
		 	 panel shows the negative log-likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors $J$ and number 
			 of clusters $K$, the middle panel shows the BIC (Equation \ref{eq:bic}), 
			 and the lower panel shows
			 the pseudo-BIC
			 (Equation~\ref{eq:pseudo-bic}). The marker indicates the 
			 lowest value in each panel.}
    \label{fig:experiment-1-gridsearch}
\end{figure}

@@

The estimated latent factor loads found during this grid search are shown in
Figure~\ref{fig:fig:experiment-1-gridsearch-factors-init-random-and-random},
where each panel shows the estimated latent factor loads found by
expectation-maximization for a given number of latent factors and components.

The latent factors $\factorloads$ in each panel are multiplied by a rotation
matrix $\mathbf{R}$ to be as similar (up to orthogonal rotation) as those found
from the same number of latent factors and $\NumComponents  =  1$.
In doing so we find that the estimated latent factors vary slightly as
the number of components $\NumComponents$ increases.

%\begin{figure*}
%	\includegraphics[width=1.0\textwidth]{experiments/toy-factors-init-randomly.png}
%	\caption{Latent factor loads inferred for models with increasing numbers
%			 of latent factors and increasing numbers of cluster components
%			 in Experiment~1, where the latent factors and component assignments
%			 were initialised randomly.
%				The top left panel shows the results of the model with
%			 $\NumLatentFactors = 1$ and $\NumComponents = 1$. 
%			 The latent factors have each been rotated by a
%			 rotation matrix such 
%			 that the estimated latent factors in each model are ordered 
%			 and oriented to be as similar as the latent factors estimated from
%			 the model with only one latent factor, up to orthogonality such
%			 that $\vec{R}\vec{R}\transpose = \eye$.
%			 }
%	\label{fig:toy-factors-init-randomly}
%\end{figure*}

%\todo{This structure goes away and becomes more consistent if we change the initialisation procedure}

%\todo{figure of toy factors init svd kmeans++}

%\todo{log-likelihood of different initialisation routines}

\begin{comment}
The results of this experiment confirm that the model is indeterminate, in 
that there are multiple equally valid solutions that can exactly reproduce
the data. It also highlights the importance of initialisation, demonstrating
how the initialisation of the common latent factors can impact the final
model parameters. These two points are related, as one could initialise the
latent factors randomly (or from noise) and multiply that model with a
rotation matrix $\mathbf{R}$ that rotates the latent space such that the
common factor loads are aligned closer to some prior expectation that we have
for the common factor loads (e.g., in that they should represent plausible
astrophysical yields). Motivated by these results and indeterminate nature
of the model, in Experiment~2 we initialise the common factor loads
to represent known astrophysical processes such that they can be interpreted
in context with the data.

@@

\todo{experiment-4-astrophysical-latent-factors}
\end{comment}

\subsection{Experiment~2: A subset of gravitationally bound stellar clusters from \Galah\ and elsewhere}
\label{sec:experiment-galah-clusters}

@@
